Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2022-10-31 analysis/paper_draft3.Rmd
updated batson figure caption and related text

A _Batson_ challenge typically proceeds in a four-step sequence. The attorney must decide whether to challenge the opposing attorney's use of one or more peremptory strikes as based on illegal bias. Second, the challenger must meet an initial burden (the _prima facie_ case) of producing just enough evidence of strike bias to raise a sufficient initial inference of illegal bias. If so, then at the next step, the striking attorney must proffer permissible reasons for their challenged strikes. Finally, the challenger must discredit those proffered reasons and ultimately persuade the trial judge to find that it is more likely than not that the striking attorney acted with illegal bias.

@@

In turn, the inference of bias based in part on historical strike data is relevant to lawyers at three of the four steps of the _Batson_ challenge, i.e., for deciding whether to argue a _Batson_ violation, for establishing a prima facie case, and for ultimately proving that attorney strike bias in the particular case is more likely than not (Figure \@ref(fig:batsonworkflowfig)).

```{r batsonworkflowfig, fig.cap="Batson workflow", fig.align = 'center', out.width="100%", fig.pos = "h"}

@@

We describe here a prototype software application ("app") that implements the approach described above and that attorneys and others can use in real time to detect bias in the use of peremptory challenges. We built this app with `r R.version$version.string` and the shiny package [@R; @shiny]. Unlike the simulations, we built the app using Rcpp [@Rcpp] and the Metropolis-Hasting algorithm to sample $b$ from the posterior distribution. To show how lawyers might use this app in real cases, we loaded this app with actual strike data from a convenience sample of attorneys who appeared during jury selection in criminal cases in the federal district court for Connecticut during fiscal years 2013 through 2017.

@@

Moreover, with a tool like our prototype app, attorneys can use the approach here in to better detect strike bias for _Batson_ and related challenges, in three ways. First, the posterior distribution of the bias parameter can help the lawyer decide in real time whether it is worth raising a  _Batson_ challenge at all.

Second, if raised, the tool can help the _Batson_ challenger meet the  evidentiary burdens of presenting a prima facie case and of ultimately proving illegal bias (Figure \@ref(fig:batsonworkflowfig)). To meet those burdens, the attorney can introduce the posterior distribution of the striking attorney's bias parameter as relevant statistical evidence. In this way, the tool helps detect bias.

But as with _any_ inference from historical strike data, the approach here in no way suffices for deciding a _Batson_ challenge. Such statistical evidence alone cannot account for the veracity of the striking attorney's proferred reasons. Besides, _Batson_ procedure allows for any type of circumstantial evidence of bias for purposes of the ultimately proving or disproving illegal bias.

At the same time, the approach here also requires more from those who would collect the appropriate historical strike data. For example, given the model we used, we required data on the cognizable-class composition of the prospective jurors not just at the time of the first strike, but at each time either party used a strike.

>>>>

A _Batson_ challenge typically proceeds in a four-step sequence. The attorney must decide whether to challenge the opposing attorney's use of one or more peremptory strikes as based on illegal bias. If challenged, the challenger must meet an initial burden (the _prima facie_ case) of producing just enough evidence of strike bias to raise a sufficient initial inference of illegal bias. If so, then at the next step, the striking attorney must proffer permissible reasons for their challenged strikes. Finally, the challenger must discredit those proffered reasons and ultimately persuade the trial judge that it is more likely than not that the striking attorney acted with illegal bias.

@@

In turn, the inference of bias based in part on historical strike data is relevant to lawyers for deciding whether to argue a _Batson_ violation, establishing a prima facie case, and proving that bias in the particular case is more likely than not (Figure \@ref(fig:batsonworkflowfig)).

```{r batsonworkflowfig, fig.cap="Inference from Historical Strike Data for  Batson Challenge", fig.align = 'center', out.width="100%", fig.pos = "h"}

@@

We describe here a prototype software application ("app") that implements the approach described above and that attorneys and others can use in real time to detect bias in the use of peremptory challenges. We built this app with `r R.version$version.string` and the shiny package [@R; @shiny]. Unlike the simulations, we built the app using Rcpp [@Rcpp] and the Metropolis-Hasting algorithm to sample $b$ from the posterior distribution. To show how lawyers might use this app in real cases in real time during selection, we loaded this app with actual strike data from a convenience sample of attorneys who appeared during jury selection in criminal cases in the federal district court for Connecticut during fiscal years 2013 through 2017.

@@

Moreover, with a tool like our prototype app, attorneys can use the approach here in to better detect strike bias for _Batson_ and related challenges. The posterior distribution of the striking attorney's bias parameter can help the lawyer decide in real time whether it is worth arguing a  _Batson_ challenge at all. If argued, the attorney can introduce that posterior distribution as relevant evidence for presenting a prima facie case and of ultimately proving illegal bias (Figure \@ref(fig:batsonworkflowfig)). To be sure, such evidence alone cannot account for the veracity of the striking attorney's proferred reasons. Besides, _Batson_ procedure allows any type of evidence, not just statistical evidence, for ultimately proving or disproving illegal bias in the particular case. Thus, as with any use of historical strike data, the approach here cannot always determine whether a _Batson_ challenge should succeed or fail.

The approach here also requires more from those who would collect the appropriate historical strike data. For example, given the model we used, we required data on the cognizable-class composition of the prospective jurors not just at the time of the first strike, but at each time either party used a strike.
